---
title: "Energetics"
output:
  pdf_document: default
  html_document: default
  
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
source("model_functions.R")
library(insol)
Lat<-median(sites$lat)
Lon<-median(sites$lon)
```

## Calculate basal metabolic rate (BMR)

This is the function taken from Humbolt paper

```{r}
FBMR <- function(mass)4.59*mass^0.69
FBMR(1.4)
mass<-(1440:1576)/1000
plot(FBMR(mass)~mass)

```


## Using the spreadsheer equation taking into account temperature and windspeed

As I understand it from the spreadsheet this set of equation should calculate the metabolic rate in the same units ()
```{r}

FMR<-function(temperature=4,windspeed=9.4,mass=1500)
{
TBrant<-37.5
DeltaT<-TBrant-temperature
b<-0.0092*mass^0.66*DeltaT^0.32
a<-4.15-b*sqrt(0.06)
a+b+sqrt(windspeed)
}
FMR()


```

## Testing against a year's climate data

I haven't got data for the whole of 2016.

```{r}
clim<-subset(clim,as.numeric(format(clim$date,'%Y'))==2015)
temperature<-(clim$tmin+clim$tmax)/20
windspeed<-0.277778*clim$avwind
plot(temperature,FMR(temperature))
```

```{r}
plot(clim$date,FMR(temperature,windspeed))
```

## Consumption function from the Humbolt paper

```{r}
biomass<-1:100
FConsumption<-function(biomass)100*0.01028*(1-exp(-0.105*biomass))*(1.0373*(1-exp(-0.0184*biomass)))
plot(FConsumption(biomass)~biomass)


```

Converting into energy assimilated

```{r}
FEnergyAssim<-function(consumption, a=0.464, E= 16.8)consumption*a*E
plot(FEnergyAssim(FConsumption(biomass))~biomass)

```

```{r}
plot(FEnergyAssim(FConsumption(sites$mean_biomass))~sites$mean_biomass)
```

#################


## A Utility function to calculate day or night

Use the insol package. This produces an object with sunrise and sunset times, so if the hour falls between them it is day.

```{r,warning=FALSE,message=FALSE}

FIsDay<-function(tm,Lat=55.32,Lon=-162.8)
{
hr<-as.numeric(format(tm, format='%H'))
day_len<-data.frame(daylength(Lat, Lon,JD(tm), tmz=-10))
isday<-ifelse(hr>day_len$sunrise & hr< day_len$sunset,"Day","Night") 
isday}

tm<-FMakeTime(2016,1,1,10)
FIsDay(tm)
tm<-FMakeTime(2016,1,1,7)
FIsDay(tm)

```

