---
title: "Energetics"
output:
  pdf_document: default
  html_document: default
  
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
source("/home/rstudio/morph/scripts/model_functions.R")
library(insol)
Lat<-median(sites$lat)
Lon<-median(sites$lon)
```

## Calculate basal metabolic rate (BMR)

This is the function taken from Humbolt paper

```{r}
FBMR <- function(mass)4.59*mass^0.69
FBMR(1.4)
mass<-(1440:1576)/1000
plot(FBMR(mass)~mass)

```


## Using the spreadsheer equation taking into account temperature and windspeed

As I understand it from the spreadsheet this set of equation should calculate the metabolic rate in the same units ()
```{r}

FBMR<-function(ftemperature=-10,fwindspeed=2,fmass=1500)
{
TBrant<-7.5
ftemperature[ftemperature>TBrant]<-TBrant
fwindspeed[fwindspeed<0.5]<-0.5
DeltaT<-TBrant-ftemperature
b<-0.0092*fmass^0.66*DeltaT^0.32
a<-4.15-b*sqrt(0.06)
a+b+sqrt(fwindspeed)
}
FBMR(10)


```

## Testing against a year's climate data

I haven't got data for the whole of 2016.

```{r}
clim<-subset(clim,as.numeric(format(clim$date,'%Y'))==2015)
temperature<-(clim$tmin+clim$tmax)/20
windspeed<-clim$avwind/10
plot(temperature,FBMR(temperature))
```

```{r}
plot(clim$date,FBMR(temperature,windspeed))
```

## Consumption function from the Humbolt paper

```{r}
biomass<-1:100
FConsumption<-function(fbiomass)100*0.01028*(1-exp(-0.105*fbiomass))*(1.0373*(1-exp(-0.0184*fbiomass))/60)
plot(FConsumption(biomass)~biomass)


```

Converting into energy assimilated

```{r}
FEnergyAssim<-function(consumption, a=0.464, E= 16.8)consumption*a*E*1000
plot(FEnergyAssim(FConsumption(biomass))~biomass)

```

```{r}
plot(FEnergyAssim(FConsumption(sites$mean_biomass))~sites$mean_biomass)
```

#################


## A Utility function to calculate day or night

Use the insol package. This produces an object with sunrise and sunset times, so if the hour falls between them it is day.

```{r,warning=FALSE,message=FALSE}

FIsDay<-function(tm,Lat=55.32,Lon=-162.8)
{
hr<-as.numeric(format(tm, format='%H'))
day_len<-data.frame(daylength(Lat, Lon,JD(tm), tmz=-10))
isday<-ifelse(hr>day_len$sunrise & hr< day_len$sunset,"Day","Night") 
isday}

tm<-FMakeTime(2016,1,1,10)
FIsDay(tm)
tm<-FMakeTime(2016,1,1,7)
FIsDay(tm)

```

## Resting


Assume 20% over BMR. Add time argument in seconds.
Returns energy used in KJoules.

```{r}

BMR<-FBMR(temperature,windspeed,fmass=1700)

FRest<-function(fbmr=BMR,ftime=3600)
{1.2*fbmr*ftime/1000}

hist(FEnergy2Fat(FRest()))
```

## Feeding

Assume use twice BMR while feeding.

```{r}
FFeed<-function(fbmr=BMR,ftime=3600,fbiomass=100){
EGain<-FEnergyAssim(FConsumption(fbiomass))*ftime/1000
EUse<-2*fbmr*ftime/1000
EGain-EUse
}
hist(FEnergy2Fat(FFeed()))
```

## Maximum energy per day

According to the Humbolt paper this is given by

```{r}
FMaxDaily<-function(mass)1713*mass^0.72
FMaxDaily(1500)/1000
FFeed()[1]
```

So using these formulae the birds can get enough energy for a whole day from just one hour's intensive feeding. This seems much too high. However this assumes that an hour feeding is completely dedicated to intensive feeding activity, which is unrealistic.

## Flying

Assume velocity of 60km per second and overhead for take off for all flights equivalent to a minute's flight time. Use 12 times BMR.



```{r}

FFlightTime<-function(fspeed=60,fdistance=1000000)
{
  speedms<-fspeed/3.6
  ftime<-fdistance/speedms+60
}
FFly<-function(fbmr=BMR,ftime=FFlightTime()){
  EUse<-12*fbmr*ftime/1000
  EUse
  }

hist(FEnergy2Fat(FFly()))
```

```{r}
save(list=ls(),file="/home/rstudio/morph/scripts/functions.rob")
```

